Liquid Amortization - Proving amortized complexity with LiquidHaskell (Functional Pearl)

### 2021-05-15 paper/main.tex
Add outlines of all sections

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2021}
\acmYear{2021}
\acmDOI{tbd}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Complexity theory for all! - Using LiquidHaskell to prove complexity of data structures (Functional Pearl)}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

\end{abstract}

@@

@@

@@

@@

@@

@@

@@

>>>>

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2021}
\acmYear{2021}
\acmDOI{tbd}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Liquid Amortization - Proving amortized complexity with LiquidHaskell (Functional Pearl)}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
TODO
\end{abstract}

@@

Motivation:
\begin{itemize}
\item{Formal proofs for properties of data structures are getting more common}
\item{Most of the time proof is done in other language}
\item{Needs a translation step, either manual or programmatic}
\item{Might have errors in translation}
\item{Might not support all features of implementation language}
\item{Thus needs encoding}
\item{Hard to reconnect proofs back to the actual implementation}
\end{itemize}

Solution:
\begin{itemize}
\item{LiquidHaskell \cite{liquidhaskell} is a refinement type \cite{refinement_types} checker for Haskell}
\item{Can be used as a theorem prover \cite{tpfa}}
\item{Directly provide proofs in Haskell, skipping the translation step}
\end{itemize}

Contributions:
\begin{itemize}
\item{Explain how amortized complexity works}
\item{How to prove amortized complexity}
\item{Show with case studies of core Haskell data structures how to apply this}
\end{itemize}

@@

\begin{itemize}
\item{Worst case is often too pessimistic}
\item{Idea: spread out expensive operations over multiple cheap ones}
\item{Usually only talks about sequence of operations without sharing}
\item{Formalization: Physicist's method}
\end{itemize}

@@

\begin{itemize}
\item{Uses refinement types}
\item{Refinments are solved by external SMT solver (z3)}
\item{With the refinement one can define proof combinators}
\item{Use equational reasoning and induction to prove interesing properties}
\end{itemize}

@@

\begin{itemize}
\item{encode requirements of potential as type synonym}
\end{itemize}

@@

\begin{itemize}
\item{definition of data structure and operations}
\item{worst case of push and multipop is O(1) and O(n)}
\item{but every item we pop was pushed at some point}
\item{define potential as height of stack}
\item{refine all functions to the logic level \cite{reflection}}
\item{guess constant factor}
\item{prove base case and adjust constant}
\item{use induction for complex case}
\end{itemize}

@@

\begin{itemize}
\item{originally described in \cite{fingertrees}}
\item{definition of finger tree}
\item{code is copy pased from the containers package}
\item{in contrast to \cite{fingertrees_new} four constructors and different append function}
\item{slightly modified because LiquidHaskell does not work great with type classes}
\item{mention polymorphic recursion that is not supported in Lean \cite{lean} and Isabelle/HOL \cite{isabelle}}
\item{potential of a finger tree}
\end{itemize}

\subsubsection{Cons and Snoc}

\begin{itemize}
\item{cons and snoc are very similar, see snoc code in appendix}
\item{definition of the cons function}
\item{definition of the consT function}
\item{step by step proof of amortized complexity}
\end{itemize}

\subsubsection{Append}\label{sec:append}

\begin{itemize}
\item{same as worst case, but potential may differ}
\item{definition of append as mutually recursive functions}
\item{no automatic termination metric, track termination across recursion}
\item{generates a lot of constraints > 3000, takes long to typecheck}
\item{defining the timing function would generate over 8000 constraints}
\item{proving with logarithms is not easy, but potentiation makes it easier}
\item{sadly not really possible to do as every step takes forever to typecheck}
\end{itemize}

@@

\subsection{Definition of the timing function}

\begin{itemize}
\item{without changing the system under test, we have to manually create a timing function}
\item{manual approach might lead to errors}
\item{Idea 1: factor out recursion and use same recursion for implementation and timing}
\item{Idea 2: use template Haskell to generate timing function}
\end{itemize}

\subsection{LiquidHaskell for verification}

\begin{itemize}
\item{support for type classes is not great at the moment}
\item{as seen in section~\ref{sec:append}, more complicated pattern matches lead to combinatorial explosion}
\item{might be possible to be smarter when generating constraints}
\item{no direct support for rationals make non-linear proofs involving division harder}
\item{several performance related options like --no-termination, --diff, --fast, did not make a difference for type checking}
\end{itemize}

\section{Conclusion}

\begin{itemize}
\item{Get more confidence that your implementation matches the theoretical expectations}
\item{Easy to apply for simple complexities like O(1) or O(n)}
\item{Harder with O(log n) because the SMT prover/LiquidHaskell is not quite there yet}
\end{itemize}
